File-based localization of numerical perturbations in data analysis pipelines

### 2020-04-01 paper.tex
First version of discussion

@@

Static vs dynamic executables

The reproducibility of a given tool may vary across subjects, and across parameters. 
See FLIRT in DC in Figure 4.
Two types of differences are observed: between-process and between-subject.
Between-process differences occur when the same tool behaves differently in different process instances.
For example, \texttt{flirt}
always produces files without differences in the \texttt{AAve} sub-pipeline but 
almost always creates different files in the \texttt{AAli} sub-pipeline.
Conversely, the same process can be reproducible for some subjects but not for others.

Propagation is very important.

  The similar findings in another study~\cite{bowring2019exploring} were reported 
  the effect of between-subject differences on reproducibility.

* granularity is that
 of processes, no in-memory checks.

Method:
* we assume no concurrent file accesses.

* we assume that the provenance graph is oriented, i.e., a process cannot read or write in the same file. Reasonnable in general but wouldn't work 
for, say, out of core computations.

Comment on the overhead, number of executions (3, 2 in condition 1, 1 in condition 2)

Applicability: no instrumentation required, fast.
But not able to capture dependencies in memory.

We identified four processes that create differences 
among all 28 distinct processes that are involved in the PreFreeSurfer execution. 
The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear 
registration processes implemented in FSL \emph{FLIRT}. 
We can see that \emph{FLIRT} creates differences in almost all the sub-pipelines. 
We monitored the library system calls using \emph{ltrace} to check that registration processes
use the same libraries. There were no differences in the library dependencies used by 
the tool in two operating systems, but we found a different number of calls to the \emph{glibc}.

In the second experiment, 
the analyses obtained by the FreeSurfer show substantial between-OS differences.
These results show the sensitivity of the segmentation process to the 
size of regions, however, it might be also sensitive to the location and shape of 
the segmented regions.
We demonstrate that static build of the pipeline would not improve the numerical stability of the pipeline. 
FreeSurfer pipeline propagates and amplifies the numerical differences that are created by the 
differences in the underlying operating system libraries.
These results are consistent with the findings in the paper~\cite{Glatard2015}, 
as it has confirmed the propagation of differences in cortical thickness extraction analysis using FreeSurfer.

We obtained results between OS versions; further investigations could be done over the OS families (e.g., Linux VS. Windows).
In addition to variations of the operating system, numerical instability of the pipelines have been 
explored when varying software, hardware and implementation details, or even by adding negligible 
perturbations like one voxel noise in the input dataset.
In the field of numerical analysis, we can investigate these issue using the Monte Carlo Arithmetic (MCA), 
a method instrumenting floating-point operations based on the random rounding algorithms at a target precision.
By this algorithm, we can locate the instabilities for the given pipelines, tools, or operations.

We demonstrate that computational models in neuroimaging are sensitive to numerical instabilities. 
Although we identify the processes in the pipeline that create differences, 
the amplification of these differences should be quantified in the future works.
We can use distance-preserving hashes rather than MD5 checksums or metrics specific to the file type like the Dice coefficient 
for the registration and segmentation processes to measure the magnitude of differences.

% mention fuzzy libmath as a way to do this across distributions.

@@

Our technique can characterize the stability of the pipelines
automatically. We identified the leading causes of the between-OS differences as (1) the evolution 
of the math libraries over time and (2) the instability of the pipelines. 
There are two ways to tackle this problem. The easy but less preferred solution is masking the instabilities.
This can be done by using, (1) single operating system for the processing of subjects, 
(2) containerizing the pipelines so that the 
processing is done on a more controlled environment, 
(3) increasing the numerical precision of the arithmetics, 
(4) building static executables by removing the host operating system library dependencies. 
These solutions only make the problem invisible, but the pipelines are still unstable.

\note{The preferred solution is to find the origin of amplification of differences and 
then fixing the particular functions/modules in the processing pipelines. 
Bagging is a fundamental technique toward reducing the instability of the analysis.
Therefore, we can use bagging for the parts in the pipeline that amplify differences, 
but it still is computationally expensive. }

% importance of data-driven evaluations, subject clustering, etc

% future work with MCA

% link registration to our motion estimation study

>>>>

\tristan{Ali, it would be useful to show a complete PreFreesurfer graph here.}

@@

\subsection{Key findings}

Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015} where fMRI pre-processing using FSL was found to be
non-reproducible across operating systems starting from the motion
correction step, a step that uses FSL's \texttt{flirt} tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it is specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is possible that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

While only a handful or processes were found non-reproducible across the
tested operating systems, including most notably linear and non-linear
registration, the effect of such instabilities were found to propagate
widely in the pipelines, resulting in Dice coefficients frequently below
0.9 for the amygdala or ventricles. This illustrates the need to conduct such 
reproducibility studies for entire pipelines rather than isolated processes.

Furthermore, as shown in Figure~\ref{fig:pfs_freq}, the reproducibility of
a given tool may vary across subjects and across processing parameters. For
instance, linear registration with FLIRT seems to be fully reproducible in
the Anatomical Average sub-pipeline, while it is highly non-reproducible in
ACPC Alignment. In Brain Extraction, the same tool was found reproducible
for some subjects only. Therefore, reproducibility studies need to be
performed on several representative subjects rather than only a couple of
them. While this is common practice in neurosciences, it is common that
software tests be executed only on a single dataset, to reduce the
associated computational load. It is likely, however, that the observed
instabilities stem from numerical edge cases that may not manifest in all
datasets.

While our findings give insight on the reasons why image analysis processes
are unstable across operating systems, they do not provide a complete
uncertainty quantification for neuroimaging pipelines across operating
systems. Although CentOS is a popular distribution in
neurosciences~\cite{hanke2011neuroscience}, different observations could be
made on other operating systems. The numerical noise introduced by
operating system updates is a realistic one, as such updates are likely to
occur throughout the time span of a neuroscience study, but it is also
uncontrolled, as it originates in updates of low-level libraries by
third-party developers. A possible method to study this problem more
comprehensively would be to introduce controlled numerical perturbations in
pipelines, which could be done either using data noise, or through
computational noise introduced by Monte-Carlo
Arithmetic~\cite{Parker1997-qq}. The work in~\cite{kiar2019comparing}
discusses and compares these two techniques.

\subsection{\toolname evaluation}

The processes identified by \toolname as non-reproducible were all
associated with dynamically-linked executables \tristan{Ali, please check
that}. This makes complete sense as statically-linked executables wouldn't
be impacted by library updates. Moreover, the hypothetical effects of
hardware or Linux kernel updates were not measured, as the different
operating sytems were deployed in Docker containers on the same host, that
is, using the same kernel and hardware.

To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead, by executing at step (2) only the
processes depending on transient files.

We demonstrated the applicability of our approach by evaluating two of the
arguably most complex pipelines in neuroimaging. Technically, these
pipelines consist of a mix of tools assembled from different toolboxes
through a variety of scripts written in different languages. Our file-based
approach, notably enabled by \reprozip, was able to analyze these pipelines
without requiring their instrumentation, which saved a very substantial
technical effort. The assumptions made on the pipeline structure, related
to the absence of concurrent writes, were not violated in our analysis, and
are likely to not impede \toolname's applicability to the most common
neuroimaging pipelines.

However, file-based analyses also have limitations related to the
granularity at which they operate. Indeed, differences can only be
identified at the level of an entire operating-system process, which can
correspond to arbitrary amounts of code. Narrowing down the analysis to
particular libraries, functions, or even code sections would require
another approach. Similarly, \toolname would not be able to detect
differences in data not saved in files, and passed to subsequent processes
in memory. A common scenario in neuroimaging pipelines is that tools return
results in their standard output, which is parsed by the calling process
and passed to subsequent ones through variables. Differences that occur in data 
that is processed in this way would not be detected by \toolname.

@@

Summary of key findings.

Spot works well but is a first step toward deeper investigation of 
particular processes.
